Plastic in the Urban Freshwater Environment

### 2021-10-29 code/analysis_code/analysis.Rmd
finishing project pt 3

#load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling

@@

# fit linear model to continuous outcome (particles_l)
```{r}
#attempt to visualize data
#look at particles per L vs population
ggplot(data, aes(x = population, y = particles_l)) + geom_point() + geom_smooth(method = lm)
```

# linear regression
```{r}
#linear regression model specification set up
lm_mod <- linear_reg() %>% set_engine("lm")
#estimating/training linear model
lm_fit1 <- lm_mod %>% fit(particles_l ~ ., data = data)
#view summary of fit
tidy(lm_fit1)
```

@@

#comparison of fit
comp2_3 <- anova(lm_fit2$fit, lm_fit3$fit, test = "Chisq")
comp2_3
```

@@

#clean up data for modelling
```{r}
data %>% is.na() %>% summary()
complete <- data %>% na.omit()
```

Trying code from above again with the complete dataset

```{r}
#estimating/training linear model
lm_fit4 <- lm_mod %>% fit(particles_l ~ ., data = complete)
#view summary of fit
tidy(lm_fit4)
#predict particles per liter based on population
lm_fit5 <- lm_mod %>% fit(particles_l ~ population, data = complete)
#view summary of fit
tidy(lm_fit5)

@@

Here, we have a VERY large RSS, which I think indicates a poor model fit for our data. Let's try with some variables removed.

# Subsetting data to remove unimportant variables
```{r}
subset <- complete %>% select(-"site", 
                              "lat", 
                              "long",
                              "date") 
#estimating/training linear model
lm_fit7 <- lm_mod %>% fit(particles_l ~ ., data = subset)
#view summary of fit
tidy(lm_fit7)

#predict particles per liter based on population
lm_fit8 <- lm_mod %>% fit(particles_l ~ population, data = subset)
#view summary of fit
tidy(lm_fit8)

#predict particles per liter based on e coli
lm_fit9 <- lm_mod %>% fit(particles_l ~ e.coli.cfu, data = subset)
#view summary of fit
tidy(lm_fit9)

@@

# recipe to fit continuous outcome to all predictors
```{r}
#recipe for particles per liter, all predictors
mp_rec <- 
  recipe(particles_l ~ ., data = train_data)

@@

# evaluate model with ROC and ROC-AUC
```{r}
#use trained workflow to predict with test data
predict(mp_fit, test_data)

@@

#trying model evaluation with SUBSET
```{r}

@@

>>>>

#load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(gridExtra) #for exporting tables

@@

# fit linear model to continuous outcome (particles_l)
```{r}
#attempt to visualize data
#look at particles per L vs population
m1 <- ggplot(data, aes(x = population, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Particles/Liter")

plot(m1)
#save figure
ggsave(filename = here("results","m_concvpop.png"), plot=m1)
```

# particles per liter vs population with outliers filtered out
```{r}
#attempt to visualize data
#look at particles per L vs population
m2 <- data %>% filter(particles_l < 750) %>% ggplot(aes(x = population, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Particles/Liter")

plot(m2)
#save figure
ggsave(filename = here("results","m_concvpop_no_outlier.png"), plot=m2)
```

# fit linear model to continuous outcome (particles_l) with log transformation
```{r}
#attempt to visualize data
#look at particles per L vs population
m3 <- ggplot(data, aes(x = population, y = particles_l)) + geom_point() +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Log(Particles/Liter)")

plot(m3)
#save figure
ggsave(filename = here("results","m_logconcvpop.png"), plot=m3)
```

# visualize particles vs e.coli.cfu 
```{r}
m4 <- ggplot(data, aes(x = e.coli.cfu, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. E. coli CFU", x = "CFU/mL", y = "Particles/Liter")

plot(m4)
#save figure
ggsave(filename = here("results","m_concvcfu.png"), plot=m4)
```

# particles per liter vs e.coli.cfu with log transformation
```{r}
m5 <- ggplot(data, aes(x = e.coli.cfu, y = particles_l)) + geom_point() + scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. E. coli CFU", x = "Log(CFU/mL)", y = "Log(Particles/Liter)")

plot(m5)
#save figure
ggsave(filename = here("results","m_logconcvcfu.png"), plot=m5)
```

# linear regression
```{r}
data <- data %>% select("particles_l", "visual_score", "turbidity.ntu", "temperature.c", "e.coli.cfu", "population", "dist")

#linear regression model specification set up
lm_mod <- linear_reg() %>% set_engine("lm")
#estimating/training linear model
lm_fit1 <- lm_mod %>% fit(particles_l ~ ., data = data)
#save table
lm_fit1_table <- tidy(lm_fit1)
#view summary of fit
(lm_fit1)
#save file
saveRDS(lm_fit1_table, file = here("results", "lm_fit_table.rds"))
```

@@

#comparison of fit
#comp2_3 <- anova(lm_fit2$fit, lm_fit3$fit, test = "Chisq")
#comp2_3
#these lines of code are kept to show how I got the original error message about models not being fitted to the same size dataset, but I have cut them out so that the code will run smoothly
```

@@

# create subset of data with only complete cases
```{r}
data %>% is.na() %>% summary() #check out missing cases
complete <- data %>% na.omit() #save complete cases
```

Trying code from above again with the complete dataset
# Modeling with complete cases only
```{r}
#estimating/training linear model
lm_fit4 <- lm_mod %>% fit(particles_l ~ ., data = complete)
#view summary of fit
tidy(lm_fit4)
#predict particles per liter based on population
lm_fit5 <- lm_mod %>% fit(particles_l ~ population, data = complete)
#view summary of fit
tidy(lm_fit5)

@@

Here, we have a VERY large RSS, which I think indicates a poor model fit for our data. Let's try with some variables removed. If an RSS of 0 is a perfect model, our RSS of 2 million is either a *very bad* model or I am making some sort of error/misinterpreting(?).

# Modeling single predictors with complete cases
```{r}
#predict particles per liter based on population
lm_fit8 <- lm_mod %>% fit(particles_l ~ population, data = complete)
#view summary of fit
tidy(lm_fit8)

#predict particles per liter based on e coli
lm_fit9 <- lm_mod %>% fit(particles_l ~ e.coli.cfu, data = complete)
#view summary of fit
tidy(lm_fit9)

@@

# recipe to fit particles/L outcome to all predictors
```{r}
#recipe for particles per liter, all predictors
mp_rec <- 
  recipe(particles_l ~ ., data = train_data)

@@

# evaluate model with RMSE
```{r}
#use trained workflow to predict with test data
predict(mp_fit, test_data)

@@

#trying model evaluation with turbidity - linear regression
```{r}
#estimating/training linear model
turb_fit1 <- lm_mod %>% fit(particles_l ~ turbidity.ntu, data = data)
#view summary of fit
tidy(turb_fit1)

#visualize linear model
t1 <- ggplot(data, aes(x = turbidity.ntu, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Turbidity", x = "Turbidity (ntu)", y = "Particles/Liter")

#save figure
ggsave(filename = here("results","m_concvturbidity.png"), plot=t1)
```

#evaluating fit using turbidity as predictor
```{r}
#recipe for particles per liter, all predictors
turb_rec <- 
  recipe(particles_l ~ turbidity.ntu, data = train_data)

#linear regression model workflow set up
turb_wflow <- 
  workflow() %>% add_model(lm_mod) %>% add_recipe(turb_rec)

#use workflow to prepare recipe and train model with predictors
turb_fit <- 
  turb_wflow %>% fit(data = train_data)

#extract model coefficient
turb_fit %>% extract_fit_parsnip() %>% tidy()

#use trained workflow to predict with test data
predict(turb_fit, test_data)

#include probabilities
turb_aug <- augment(turb_fit, test_data)

#get RMSE for test data
turb_aug %>% rmse(truth = particles_l, .pred)

@@

Here, I am still confused about why so many rmse values are resulting. Earlier, I was getting multiple rmse values but thought it was because I was building a model with all predictors. Here, I am using one predictor but still getting multiple rmse values. If I remember correctly, there was only one rmse value in the exercises we worked on in the previous weeks.

### 2021-10-29 code/analysis_code/exploration.Rmd
finishing project pt 3

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)
library(geodist) #to calculate distance from sample site to each water reclamation facility
library(GGally) #for producing correlation matrices

@@

>>>>

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)
library(GGally) #for producing correlation matrices

@@

# Summary statistics
```{r}
summary <- summary(mpdata)
summary
#save table
saveRDS(summary, file = here("results", "summarytable.rds"))
```

### 2021-10-29 code/processing_code/censusland_code.Rmd
fixing errors, building up analysis portion
finishing project pt 3

##Load Dependencies
The following code loads the package dependencies for our analysis:
```{r}
#install.packages("tidycensus", repos = "http://cran.us.r-project.org")
#install.packages("censusapi", repos = "http://cran.us.r-project.org")
#install.packages("censusGeography", repos = "https://cran.rstudio.com/bin/macosx/contrib/4.0/chron_2.3-55.tgz")
#install.packages("rerddap", repos = "https://cran.rstudio.com/bin/macosx/contrib/4.0/rerddap_0.6.5.tgz")

@@

#path to data
data_location <- here::here("data","processed_data","cleansampledata.rds")

#load data. 
data <- readRDS(data_location)
```

@@

>>>>

##Load Dependencies
The following code loads the package dependencies for our analysis:
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
#install.packages("tidycensus", repos = "http://cran.us.r-project.org")
#install.packages("censusapi", repos = "http://cran.us.r-project.org")
#install.packages("censusGeography", repos = "https://cran.rstudio.com/bin/macosx/contrib/4.0/chron_2.3-55.tgz")
#install.packages("rerddap", repos = "https://cran.rstudio.com/bin/macosx/contrib/4.0/rerddap_0.6.5.tgz")

@@

```

@@

#cleanup resultant dataset
athensland <- athensland %>% select(-"fips", 
                              -"FIPS", 
                              -"AREA", 
                              -"quarter", 
                              -"month",
                              -"year")

### 2021-10-29 code/processing_code/censusprocessing.Rmd
various edits
update exploration figures, manuscript additions, adding data index and populating readme files
finishing project pt 3

saveRDS(fips, "fips.RDS")
```

@@

# Using NHGIS data for more recent block-level population counts
```{r}
#Downloaded data from http://NHGIS.org
install.packages("ipumsr")
library(ipumsr)

>>>>

saveRDS(fips, here("data", "processed_data", "fips.RDS"))
```

@@

# Using NHGIS data for more recent block-level population counts
```{r}
#Downloaded data from http://NHGIS.org
library(ipumsr)

### 2021-10-29 code/processing_code/wrfprocessing.Rmd
finishing project pt 3

I also wanted to look at proximity of sample sites to water reclamation facilities (WRFs). Let's try that. 
# load WRF location
```{r}
#load MP data
mpdata <- readRDS(here("data", "processed_data", "popdata.rds"))
#path to data
wrfdata_location <- here::here("data","processed_data","wrflocation.rds")

>>>>

I also wanted to look at proximity of sample sites to water reclamation facilities (WRFs). Let's try that. 
# load WRF location
```{r}
library(here)
library(geodist) #to calculate distance from sample site to each water reclamation facility
library(dplyr) #for data processing
library(tidyverse)

#load MP data
mpdata <- readRDS(here("data", "processed_data", "popdata.rds"))
#path to data
wrfdata_location <- here::here("data","processed_data","wrflocation.rds")

### 2021-10-29 products/manuscript/Manuscript.Rmd
finishing project pt 3

Additional data [write up further explanation]:
 - small area population via census zipcode
 - UOWN bacteria level data
 - local land cover
 - distance from water reclamation facility (WRF)

@@

My ideas for how to analyze this data include using non-parametric statistical tests, correlation matrices, and modeling (?) to characterize and compare observations.

@@

- Methods for sample collection and processing for UOWN MP water samples

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

@@

Import of additional data
 - census - American Community Survey
 - NGIS land cover
 - zip code - Bing reverse geocoding
 - WRF locations: located on Google Maps, coordinates recorded in Excel spreadsheet
 - UOWN bacteria data (citation/attribution available on UOWN.org)

@@

This section will contain the key products of my exploratory analysis (located in exploration.Rmd) when complete. Since I am keeping this document in manuscript style, I will keep processing/exploration/analysis code separate. 
_Use a combination of text/tables/figures to explore and describe your data. You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. FIgures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data._

_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones._

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

Table \@ref(tab:summarytable) shows a table summarizing the data.

```{r summarytable,  echo=FALSE}
#saving template for formatting later on

#resulttable=readRDS("../../results/summarytable.rds")
#knitr::kable(resulttable, caption = 'Data summary table.')
```

Figure \@ref(fig:resultfigure) shows a scatterplot figure produced by one of the R scripts.

```{r resultfigure,  fig.cap='Analysis figure.', echo=FALSE}
#knitr::include_graphics("../../results/resultfigure.png")
```

@@

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.

```{r resulttable,  echo=FALSE}
#resulttable=readRDS("../../results/resulttable.rds")
#knitr::kable(resulttable, caption = 'Linear model fit table.')
```

>>>>

@@

@@

### Description of study area
The Upper Oconee Watershed is located in Northeast Georgia, spanning from Gainesville on the northern end to Lake Sinclair in Milledgeville, Georgia, on the southern end. The Upper Oconee Watershed encompasses 7590 km2, and includes more than 4000 km of streams [@Fisher]. Land within the Upper Oconee Watershed is historically agricultural, but has been subject to increasing urbanization in the past two decades; the study area contains developed and densely populated urban residential areas [@Cho; @Fisher]. The Oconee Watershed is one of 14 river basins in the state of Georgia; the Oconee River feeds into the Altamaha River, which eventually deposits into the Atlantic Ocean off the Georgia coast (source: UOWN website). The Oconee River begins where the North and Middle Oconee Rivers meet, south of Athens, Georgia. The North and Middle Oconee Rivers begin in Lula, GA, and Brassleton, GA, respectively.

The Upper Oconee Watershed is regularly monitored by the Upper Oconee Watershed Network (UOWN), an Athens-based monitoring, education, advocacy, and recreation organization. Quarterly sampling events occur throughout the year, and the present research project was made possible via collaboration and eventual partnership with UOWN and the quarterly sampling events. The UOWN study area consists of 600 km2.

### Experimental design: sample collection and processing
1-liter grab samples of water were collected quarterly from Fall 2020 through Summer 2021, in November, February, April, and July by UOWN community members. Samples were collected from 25 to 30 sites on each sampling date. The number of sites was decreased from previous sampling events due to COVID-19 public health restrictions with volunteers.

Samples were processed via vacuum filtration, which was performed in 150 mL volumes in duplicate for each site. Samples were pumped through gridded 0.45 micron membranes (Whatman) and immediately covered and stored covered in plastic petri dishes at room temperature for up to 8 weeks. 
Filters were visually inspected using a dissecting microscope with magnification ranging from [INSERT]. Microplastic particles were identified based on the following characteristic features: unnatural homogenous color, unnatural shape, homogenous texture, homogenous width (no tapered end) (CITE). Microplastic concentration (No/L) was calculated by dividing the total microplastic particle count by the sample volume. Duplicates (samples A and B for each site) were averaged to obtain a singular concentration value for each sample site.

### Controls
To evaluate contamination in the lab, blank filters were used to gauge airborne contamination and contamination within processing methods. Blank filters were placed in uncovered petri dishes on each work bench during sample filtration and processing. DI and MilliQ water were filtered in 150 mL volumes, in duplicate, to produce controls to represent potential contamination from filtration processes. The average contamination per sample was = INSERT.

## Data import and cleaning

@@

To supplement my analysis, additional data was imported as detailed below.

### UOWN Quarterly Data
Public data from Upper Oconee Watershed Network, Athens, Georgia; retrieved from www.uown.org
### Population
NHGIS [@Mansen]
### Land Cover
NHGIS [@Mansen]
### Water Reclamation Facilities
Coordinates for Athens-Clarke county's three water reclamation facilities: North Oconee, Middle Oconee, and Cedar Creek were manually located via Google Maps.

@@

This section will contain the key products of my exploratory analysis (located in exploration.Rmd) when complete. Since I am keeping this document in manuscript style, I will keep processing/exploration/analysis code separate.

Table \@ref(tab:summarytable) shows a table summarizing the data.
```{r summarytable,  echo=FALSE}

#summarytable=readRDS("../../results/summarytable.rds")
#knitr::kable(summarytable, caption = 'Data summary table.')

#I eventually want a nice looking summary statistic table. This is a place holder for now  
```

### Distribution of Microplastic Concentration

Figure \@ref(fig:distribution) shows a histogram of microplastic concentration observations. The minimum concentration is 16.67 particles/L and the maximum is 1193.33 particles/L. The mean concentration is 104.39 particles/L, and the median is 66.67 particles/L.

```{r distribution,  fig.cap='Distribution of Microplastic Concentration', echo=FALSE}
knitr::include_graphics("../../results/distribution.png")
```

Microplastic concentrations remained in similar ranges throughout the study period. Figure \@ref(fig:concentrationbydate) shows a boxplot of concentrations by sample date.

```{r concentrationbydate,  fig.cap='Particles/L by Sample Date', echo=FALSE}
knitr::include_graphics("../../results/concentrationbydate.png")
```

There is some seasonal variation in concentration at each individual site. Figure \@ref(fig:seasonalconc-bysite) shows a plot of concentrations at each site.

```{r seasonalconc-bysite,  fig.cap='Seasonal Variation in Particles/L', echo=FALSE}
knitr::include_graphics("../../results/seasonalconc_bysite.png")
```

There are similar microplastic levels throughout the watersheds within the Upper Oconee. Some watersheds experienced greater variation in microplastic levels than other watersheds. Figure \@ref(fig:conc-bywatershed) shows the microplastic concentrations by watershed.

```{r conc-bywatershed,  fig.cap='Watershed Microplastic Concentrations', echo=FALSE}
knitr::include_graphics("../../results/conc_bywatershed.png")
```

Figure \@ref(fig:watershed-overtime) shows a line graph of the mean watershed microplastic concentrations at each seasonal sampling date.

```{r watershed-overtime,  fig.cap='Microplastic Concentration Over Time', echo=FALSE}
knitr::include_graphics("../../results/watershed_overtime.png")
```

### Predictors
Population, land cover/use, and bacteria levels are hypothesized predictors of microplastic concentration. Figure \@ref(fig:particlesvpop) and Figure \@ref(fig:logmpvcfu) demonstrate the relationship between microplastic concentration and population and microplastic concentration and bacteria levels (CFU/L), respectively.

```{r particlesvpop,  fig.cap="Particles/L vs Population", echo=FALSE}
knitr::include_graphics("../../results/particlesvpop.png")
```

```{r logmpvcfu,  fig.cap='Log particles/L vs CFU', echo=FALSE}
knitr::include_graphics("../../results/logmpvcfu.png")
```

Figure \@ref(fig:matrix) and Figure \@ref(fig:landmatrix) show correlation matrices for the hypothesized predictor and for the different categories of land use.

```{r matrix,  fig.cap='Predictor matrix', echo=FALSE}
knitr::include_graphics("../../results/matrix.png")
```

```{r landmatrix,  fig.cap='Land cover matrix', echo=FALSE}
knitr::include_graphics("../../results/landmatrix.png")
```

@@

Preliminary modeling reveals that there is not a strong relationship between microplastic concentration and population level. Figure \@ref(fig:m-concvpop-no-outlier) demonstrates a linear model fit.

```{r m-concvpop-no-outlier,  fig.cap='Concentration vs Population Linear Model', echo=FALSE}
knitr::include_graphics("../../results/m_concvpop_no_outlier.png")
```

Figure \@ref(fig:m-logconcvcfu) shows a linear model of microplastic concentration vs CFU (both variables log-transformed).

```{r m-logconcvcfu,  fig.cap='Concentration vs CFU Linear Model', echo=FALSE}
knitr::include_graphics("../../results/m_logconcvcfu.png")
```

Figure \@ref(fig:concvturbidity) demonstrates a linear model of particles/L vs turbidity.

```{r m_concvturbidity,  fig.cap='Concentration vs Turbidity Linear Model', echo=FALSE}
knitr::include_graphics("../../results/m_concvturbidity.png")
```

Table \@ref(tab:lmfittable) shows a table summarizing a linear model fit predicting particles/L with 6 predictors.

```{r lmfittable,  echo=FALSE}
lm_fit_table=readRDS("../../results/lm_fit_table.rds")
knitr::kable(lm_fit_table, caption = 'Linear model fit table.')
```
